---
title: "Data Types and Data Structures"
format:
  html:
    toc: false
---

# Basic data types

R has the following basic data types:

-   **Numeric** to represent numbers that may contain decimal points
-   **Integer** to represent integer (whole number) values
-   **Complex** to represent numbers with both a real and imaginary part: for example, $5 + 3i$ where $i = \sqrt{-1}$
-   **Logical** to represent `TRUE` and `FALSE` 
-   **Character** to represent text data: for example, `"abc"` or `'123'`.  Note that the single or double quotes aren't part of the value, but they are used to enclose character values to ensure they are interpreted as data rather than code, and as text rather than numbers.

# Vectors

A vector is an ordered collection of data values that are all the same data type.

We can use the `c()` ("combine") function in R to combine a sequence of values into a vector. For example:

```{r}
vector1 <- c(5, 5, 9)
vector1
```

By default, the index of each item is its numbered position (i.e.). But if we want, we can also label each position as well:

```{r}
vector2 <- c(a = 5, b = 5, c = 9)
vector2
```

Remember that vectors contain values that are *all the same type*. What happens if we try to combine differently-typed values into a vector using `c()`?

# Matrices

A matrix is a 2-dimensional structure that contains data that are all the same data type.

We can use the `matrix()` function in R to create a matrix from a sequence of values. For example:

```{r}
matrix1 <- matrix(c(5, 5, 9, 7, NA, 3), nrow = 3, ncol = 2)
matrix1
```

# Lists
